Is it fair? Searching for standards of fairness in the transportation justice literature

### 2024-05-10 Journal-manuscript/Journal-manuscript.qmd
Verision for AP review

---
title: "Searching for standards of fairness in the transportation justice literature"
format: docx

#   tandf-pdf:
#     keep-tex: true
#   tandf-html: default
# author:
#   - name: Anastasia Soukhov
#     affiliations:
#       - ref: MAC
#     orcid: 0000-0003-2402-304X
#     email: soukho@mcmaster.ca
#   - name: Ignacio Tiznado-Aitken
#     affiliations:
#       - ref: TORONTO
#     orcid: 0000-0003-2402-304X
#     email: i.tiznadoaitken@utoronto.ca
#   - name: Matthew Palm
#     affiliations:
#       - ref: UNC
#     orcid: 0000-0003-2402-304X
#     email: palmmatt@unc.edu
#   - name: Steven Farber
#     affiliations:
#       - ref: TORONTO
#     orcid: 0000-0003-2402-304X
#     email: steven.farber@utoronto.ca
#   - name: Antonio Páez
#     affiliations:
#       - ref: MAC
#     orcid: 0000-0003-2402-304X
#     email: paezha@mcmaster.ca
# 
# affiliations:
#   - id: MAC
#     name: McMaster University
#     department: School of Earth, Environment and Society
#     city: Hamilton, ON
#     country: Canada
#   - id: TORONTO
#     name: University of Toronto Scarborough
#     department: Department of Geography and Planning
#     city: Toronto, ON
#     country: Canada
#   - id: UNC
#     name: The University of North Carolina at Chapel Hill
#     department: Department of City and Regional Planning
#     city: Chapel Hill, NC
#     country: USA
# keywords:
  # - transportation
  # - equity
  # - literature review

@@

```{r, eval=FALSE}
Pflow_chart |>
  DiagrammeRsvg::export_svg() |> 
  charToRaw() |> rsvg::rsvg_png(file="figures/Pflow_chart.png")
```

@@

df_all_sub_ob_just_reordered$ID = seq.int(nrow(df_all_sub_ob_just))
```

```{r}
df_all_sub_ob_just_reordered_plot <- ggplot(data = df_all_sub_ob_just_reordered, 
       aes(y = reorder(sub_ob_j_type, count),
           x = count, 
           label = sub_ob_j_type,
           fill = sub_ob_j_type))+ 
  geom_bar(stat="identity",  col = "grey", alpha=0.8) +
  facet_grid(rows=vars(sub_ob_j),
             scales="free_y",
             space="free_y",
             switch = "y") +
   scale_x_continuous(labels=scales::percent,
                      breaks = seq(0,0.8, by=0.2),
                      limits = c(0,0.65),
                      expand = c(0,0))+
  scale_fill_manual(values=c(rep(RColorBrewer::brewer.pal(9,"Pastel1"),4))) +
  geom_text(size = 2.5, position = position_stack(vjust = 1), col = "Black") +
  theme_classic() +
  theme(strip.text.y = element_text(size = 5.5),
        strip.background.y = element_rect(fill = "lightgrey"),
        strip.placement = "inside",
        axis.title = element_blank(),
        axis.ticks.y=element_blank(),
        axis.text.y=element_blank(),
        axis.text.x=element_text(size=5),
        legend.position = "none",
        panel.grid.major.x = element_line(colour="grey", linewidth=0.1))
```

@@

```{r}
standards_conc_percentages_plot <- ggplot(totals_standards_conc, 
       aes(x=std_mcat, y=conc, fill= total_perc_in_all ))+ 
  geom_tile() + 
  scale_fill_gradient2(low = "#AFAFED", mid= "#3F3FD4", high = "#191970", midpoint = 0.1, limits = c(0,.21), na.value = "grey40") +
  scale_x_discrete(limits=(totals_standards_conc$std_mcat)[order(totals_standards_conc$group)][c(4,3,2,1)]) + 
  scale_y_discrete(limits=(totals_standards_conc$conc)[order(totals_standards_conc$group)] ) +
  geom_text(aes(label = total_perc_in_all |> percent(accuracy = 1)), color = "white", size = 2.5) +
  theme_classic() +
  coord_flip()+
  theme(legend.position = "none",
        axis.ticks = element_blank(),
        axis.text.x=element_text(angle=20,hjust=1, colour = "black"),
        axis.text = element_text(size=7, colour = "black"),
        axis.title = element_blank())

#standards_conc_percentages_plot
# ggsave("figures/standards_conc_percentages_plot.png")
```

@@

```{r, warning=FALSE}
#the heatmap on top and table of stand/concept definitions on the bottom.
standards_conc_percentages_plot_and_table <- cowplot::plot_grid(
  gen_grob(totals_standards_table_flex, scaling="fixed", fit="auto"),
  standards_conc_percentages_plot,
  NULL,
  gen_grob(totals_conc_table_flex, scaling = "fixed", fit="auto"),
  nrow = 2, ncol=2, rel_heights = c(0.8,0.7), rel_widths = c(0.8,4) )
# standards_conc_percentages_plot_and_table
# ggsave("figures/standards_conc_percentages_plot_and_table_new.png")
```

@@

Transportation systems, as a class of essential technologies that facilitate or impede social inclusion and activity participation <!---[@churchTransport2000; @lucasTransport2001; @socialExclusion2003; @cassSocial2005; @casasSocial2007; @prestonAccessibility2007; @paezMobility2009]-->, have increasingly come into focus from the perspective of equity; a lively and rapidly growing literature has emerged on the topic [@martensTransport2016; @karnerAdvancesPitfallsMeasuring2023; @vecchioTransportEquityLatin2020]. <!--@diciommoTransport2017; @guoSystematicOverviewTransportation2020; @karnerFrom2020; @pereiraTransportationEquity2021; @vanweeEvaluating2021; @zhangLiterature2021; @desjardinsExaminingEquityAccessibility2022; --> Aside from an ethical motivation, keeping track of objective and perceived inequalities could be of interest for governing bodies to respond to popular demands for fairness. However, transportation systems are notoriously complex: they simultaneously move people, goods, and information. Emerging technologies and service models can swiftly change the balance of benefits and burdens among a population [@guoSystematicOverviewTransportation2020], turning both users and service providers into digital rentiers [@birchBig2022]. The benefits and burdens of transportation systems are diffuse over space and time posing a challenge in evaluating and tracking inequities. Over space transportation systems engineered to offer higher mobility for people _somewhere_, can simultaneously cut others off from essential opportunities _elsewhere_, as @rajeEngineering2004 poignantly illustrated with examples of infrastructure in the UK. Over time, the shades of policies past can continue to haunt a region and the planet as shown by the legacy of displacement and decay caused by urban highways all across the US [@archerWhite2020] and the time horizon for the impacts of climate change [@markolfTransportation2019].

<!--But complexity is no excuse to shirk away from the task.--> The objective of this work is to scan the state of knowledge in terms of defining and operationalizing "fairness" in the transportation domain. Much research has been devoted to the issues of _measuring_ equity in transportation, including (among many others) [@ramjerdiEquity2006; @delboscUsing2011; @welchMeasure2013; @martensMeasuring2019; @pritchardNew2022]. Further, there are multiple works that discuss the conceptual and philosophical foundations of equity and fairness in transportation [@martensTransport2016; @pereiraDistributiveJusticeEquity2017a; @vanoutriveHow2019]. There have also been, previous reviews of planning documents have investigated equity from narrowly scoped perspectives, such as accessibility [@boisjolyHowGetThere2017] or a particular mode of transportation e.g., cycling [@doranPursuitCyclingEquity2021]. These efforts are valuable to scholars, the public, and decision-makers alike. In our estimation however, there remains a gap in the literature in terms of understanding how standards for equity are developed and implemented within the transportation domain.

To illustrate this gap, we note how @beilerExploring2016 in their exploration of transport equity, cite the following strategies identified by the US DOT to address matters of justice [p. 287]:

@@

While commendable, the strategies are too vague, which means it is possible to implement them in a myriad ways, either genuinely to comply with the spirit of justice, or else performatively to deceive it [@mcculloughPerformative2023]. Some relevant questions include: how much should the adverse effects be reduced? To zero? Or to some tolerable level of adversity greater than zero? What should that level be and by what criteria? What benefits should be distributed? Should the benefits be based on simple population weights or, contrariwise, advantage more deprived individuals? These questions boil down to the development and use of _standards_ for transportation justice. In this work, we choose to focus exclusively on the academic domain as we know that adoption of equity concepts in planning practice has lagged developments in academic work [@pereiraTransportationEquity2021; @boisjolyHowGetThere2017; @linovskiEquityTransportationPlanning2020; @doranPursuitCyclingEquity2021; @litmanEvaluatingTransportationEquity2022]

Supporting the creation of (more) just transportation systems involves understanding the production and management of transportation benefits and costs; how they are distributed and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021]. In this spirit, our work seeks to make two significant contributions. First, it outlines a conceptual and flexible framework with supporting definitions for structuring transportation fairness and equity analysis. Second, it applies the framework to collate the existing academic knowledge that undertakes transportation equity analysis supported by equity standards and conceptualisations. We hope the findings and the presentation of academic knowledge will be useful in supporting the development and implementation of equity standards for transportation planners, policy and decision-makers.

In contrast to previous reviews on how to measure inequality in transportation systems, this work is concerned with the implied or explicit standards that are used to judge whether the measured inequalities are fundamentally “fair” or unacceptable. The present study achieves this objective by scanning the state of academic knowledge in defining “fairness” in the transport domain. Borrowing from philosophers of justice [@jaggarPhilosophicalChallengesGlobal2009], we ask the following questions to the literature and use them as a framework for presentation:

-	**What** are those benefits and burdens of transportation systems? Are they qualified in sidewalk widths, bus frequency, travel times, accessibility, or some other consideration?  
-	**Who** benefits and is burdened by transportation systems? Along which lines of identity are transportation inequities examined?
-	**Where and when** are equity standards applied? What are the differences between the research depending on location, context-specific planning processes and time of publication? 
-	**How** are equity standards and inequities measured? Are they infrastructure provision thresholds, acceptable pollution levels, or some other kind of cut off? How are these studies born out of broader notions of justice, i.e. disability rights, environmental justice, egalitarianism, or something else?

<!-- The structure of this review is as follows. After this introduction, the stage for our investigation is set by laying out some important definitions and describe the methods used for searching, selecting, and reviewing the relevant literature. This is followed by a summary of the findings and their critical appraisal. We conclude with some calls for action to improve the practice of setting and using standards of fairness for transportation justice. -->

@@

In the introductory section the terms "justice", "fairness", and "equity" are used relatively loosely. This was done purposefully. As noted, people often have strong intuitions of what is "fair", "just", and "equitable" which may or may not match those of the authorities who set the standards of fairness. But in democratic societies, the authority of all those charged with the business of governing should derive from the will of the people. It is therefore important to explicitly state how we plan to use these words - what we mean by "justice", "fairness", and "equity".

@@

- **Retributive justice** is concerned with the proportional retribution of wrongdoers with relation to legitimate punishers and the innocent [@sep-justice-retributive].
- **Reparative (or restorative) justice** focuses on the reparation of caused harm; it centers 'reintegrative shaming' to restore victims, wrongdoers, and the community according to moral values [@tylerRestorativeJusticeProcedural2006].
- **Procedural justice** strives to ensure that the views and preferences of all stakeholders are fairly accounted for in the decision-making and inter-personal procedures affecting their lives and communities [@tylerRestorativeJusticeProcedural2006].
- **Distributive justice** is perhaps the most commonly studied form of justice [see @jaggarPhilosophicalChallengesGlobal2009, p. 2; @pereiraDistributiveJusticeEquity2017a], and its main concern is the way the benefits and burdens of the tangible and intangible products of society are collected by different segments of a population.

@@

**Equity**, as conceptualized alongside distributive justice, tends to encompass tools to understand the distribution of benefits and burdens of things among a population; often emphasizing those with the least advantage and equality in their outcomes. In the transportation domain, the term is somewhat loaded because it is perceived as stemming from the authority of the state and is meant to assist with decisions about regulating and financing transportation spending [@karnerTransportationEquityTransportation2020]. Here, we are in agreement with @karnerTransportationEquityTransportation2020 that equity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual and perceived inequities. In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in the "political ecology of the estimated truth" [@kingModels1993]: in principle their assumptions and scope must be open and transparent, or else they may be more vulnerable to misuse and even abuse as tools of subjugation.

@@

**Standards** are a way of making concrete statements about fairness. Returning to the ambiguities in Title VI discussed by @martensFair2021, the attainment of justice depends on the standard used to indicate fairness. For example, explicit non-discrimination constitutes a very weak standard that takes aim at the actions of agencies instead of the recipients of the benefits. Accordingly, any distribution of benefits would be considered fair, as long as the agency does not explicitly and knowingly target or deny benefits to particular groups. The standard provides conditions to determine whether a situation is _fair_. A similarly weak standard is a _Pareto improvement_, whereby it is possible to concentrate the benefits as long as no group is worse off compared to the status quo. A policy is fair as long it does no harm. A somewhat more strict standard, a Pareto-Plus improvement, stipulates that an intervention is fair when all groups receive at least _some_ (non-trivial) benefits; the size of the benefits for each group is irrelevant. In contrast to the notion of "do no harm", such a standard embodies the ideal that no one is denied benefits. An egalitarian standard would weigh the benefits or burdens by population, and fairness is achieved when each group gives or receives in proportion to their size. In contrast, an affirmative action standard is even stricter, since it requires the benefits to be distributed in a non-egalitarian way that favors those who are still harmed by past or present discriminatory practices.

@@

In the case of justice in transportation, the question of **"Where?"** is paramount, as it might be argued that transportation generates inequalities by their very nature. By concentrating the effects of space-time convergence (for instance, by providing access to a transit system or a highway), an inequality is automatically generated. The burdens of transportation, in contrast, are often diffuse. They are incrementally paid, for example by a distributed population in the form of taxes, or by a population with a different spatial distribution in the form of poor health. As such, the answer to "Where?" is the definition of the spatial boundaries.

Conventionally, the question of **"When?"** refers to the temporal circumstances within which the demands of justice have application. In the cause of transportation justice, we ask about the temporal aspects of transportation systems, as examples: _when_ did the equity analysis take place and under what historical policy context; _the right time_ to invest in transportation infrastructure and as a result when to generate a spatial inequality [e.g., @rabelloEvaluation2015]; for _how long_ the burdens and benefits can still be associated to a specific transportation intervention; or even _timelines_ of reparative justice interventions that reconcile the shadows of past transportation-related injustices.

When answering **"Who?"**, we inquire about which entities should be regarded as subjects/arbiters of justice, meaning those entitled to make claims of moral consideration from the perspective of justice. To make it tractable, this question is often approached through the filter of population groups, which may include several concurrent traits, such as gender identity, ableness, ethnicity, age, caste, and income. Often, it is appropriate to reflect on the intersections between traits, given differences in a person's lived experiences. A possible complication in the case of transportation is that disentangling the "Who" from their mobility tools is not always straightforward. Although a person is not their mode of transportation, there are large segments of the population who live in situations where they cannot extricate themselves from the mobility tools they can use, either because they have driven themselves out of choices [@laveryDriving2013], or have been driven out of choices by factors beyond their control [e.g., captive users of a single mode @jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018]. <!-- In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume.-->While it is important to avoid conflating the "Who" with the "What", for analytical purposes we need to be mindful of the connection between aperson and their mobility tools. <!-- In the case of transportation, in addition to members of the public who use transportation systems, there is another category of **Who**, that stands possibly in opposition to users, namely the entities charged with providing services, maintaining infrastructure, and so on. These could be ministries or departments of transportation, transit agencies, public works departments and others having the power to act upon transportation (in)justices. Identifying these entities is relevant to elucidate who is responsible, for example for apportioning the benefits or mitigating the burdens of transportation.-->

>>>>

---
title: "Is it fair? Searching for standards of fairness in the transportation justice literature"
format: docx
csl: apa7.csl
#   tandf-pdf:
#     keep-tex: true
#   tandf-html: default
# author:
#   - name: Anastasia Soukhov
#     affiliations:
#       - ref: MAC
#     orcid: 0000-0003-2402-304X
#     email: soukho@mcmaster.ca
#   - name: Ignacio Tiznado-Aitken
#     affiliations:
#       - ref: TORONTO
#     orcid: 0000-0003-2402-304X
#     email: i.tiznadoaitken@utoronto.ca
#   - name: Matthew Palm
#     affiliations:
#       - ref: UNC
#     orcid: 0000-0003-2402-304X
#     email: palmmatt@unc.edu
#   - name: Steven Farber
#     affiliations:
#       - ref: TORONTO
#     orcid: 0000-0003-2402-304X
#     email: steven.farber@utoronto.ca
#   - name: Antonio Páez
#     affiliations:
#       - ref: MAC
#     orcid: 0000-0003-2402-304X
#     email: paezha@mcmaster.ca
# 
# affiliations:
#   - id: MAC
#     name: McMaster University
#     department: School of Earth, Environment and Society
#     city: Hamilton, ON
#     country: Canada
#   - id: TORONTO
#     name: University of Toronto Scarborough
#     department: Department of Geography and Planning
#     city: Toronto, ON
#     country: Canada
#   - id: UNC
#     name: The University of North Carolina at Chapel Hill
#     department: Department of City and Regional Planning
#     city: Chapel Hill, NC
#     country: USA
# keywords:
  # - transportation
  # - equity
  # - literature review

@@

```{r, eval=FALSE}
Pflow_chart |>
  DiagrammeRsvg::export_svg() |> 
  charToRaw() |> rsvg::rsvg_png(file="figures/Pflow_chart.tiff")
```

@@

df_all_sub_ob_just_reordered$ID = seq.int(nrow(df_all_sub_ob_just))

#adding color column 
df_all_sub_ob_just_reordered <- df_all_sub_ob_just_reordered |>
  mutate(colour = c("#FCBBA1", "#FC9272", "#FB6A4A", "#EF3B2C",
                     "#D9D9D9", "#BDBDBD",
                    "#E5F5E0", "#C7E9C0", "#A1D99B", "#74C476", "#41AB5D", "#238B45",
                    "#F7FBFF", "#DEEBF7", "#C6DBEF", "#9ECAE1", "#6BAED6", "#4292C6", "#2171B5", "#08519C", "#08306B","#08519C",
                    "#FCFBFD", "#EFEDF5", "#DADAEB", "#BCBDDC", "#9E9AC8", "#807DBA", "#6A51A3", "#54278F"))
#c(RColorBrewer::brewer.pal(9,"Blues"))
```

```{r, echo=FALSE}
df_all_sub_ob_just_reordered_plot <- ggplot(data = df_all_sub_ob_just_reordered, 
       aes(y = reorder(sub_ob_j_type, count),
           x = count, 
           label = sub_ob_j_type,
           fill = sub_ob_j_type))+ 
  geom_bar(stat="identity",  col = "grey", alpha=0.8, size=0.05) +
  facet_grid(rows=vars(sub_ob_j),
             scales="free_y",
             space="free_y",
             switch = "y") +
   scale_x_continuous(labels=scales::percent,
                      breaks = seq(0,0.8, by=0.2),
                      limits = c(0,0.65),
                      expand = c(0,0))+
  scale_fill_manual(values= c("1992 to 2007" = "#FCBBA1",
                              "2008 to 2012" =  "#FC9272", 
                              "2013 to 2017" =  "#FB6A4A",
                              "2017 to Mar. 2021" =  "#EF3B2C",
                              "Global North" =  "#D9D9D9",
                              "Global South" =  "#BDBDBD",
                              "(Dis)Abilities" =  "#DEEBF7", 
                              "Age" =  "#6BAED6", 
                              "All populations" =  "#F7FBFF",
                              "Comp. vul" =  "#C6DBEF", 
                              "Education level" =  "#6BAED6", 
                              "Employment 3.a" =  "#9ECAE1", 
                              "Gender" =  "#DEEBF7", 
                              "Income" =  "#4292C6", 
                              "Other 3.a" =  "#9ECAE1",
                              "Race/Ethnicity" =  "#C6DBEF",
                              "Bicycle" =  "#A1D99B", 
                              "Car" =  "#74C476", 
                              "Other 3.b" =  "#E5F5E0", 
                              "Pedestrian" =  "#41AB5D",
                              "Transit" =  "#238B45",
                              "Unspecified 3.b" =  "#C7E9C0",
                              "Childcare" =  "#BCBDDC", 
                              "Community" =  "#DADAEB", 
                              "Education" =  "#EFEDF5", 
                              "Employment 4." =  "#9E9AC8", 
                              "Healthcare" =  "#807DBA", 
                              "Leisure" =  "#DADAEB", 
                              "Shops" =  "#BCBDDC", 
                              "Unspecified 4." =  "#6A51A3")) + 
                      #c(rep(RColorBrewer::brewer.pal(9,"Pastel2"),4))) +
  geom_text(size = 2.5, position = position_stack(vjust = 1), col = "Black") +
  theme_classic() +
  theme(strip.text.y = element_text(size = 5),
        strip.background.y = element_rect(fill = "lightgrey", linewidth=0.3),
        strip.placement = "inside",
        axis.title = element_blank(),
        axis.ticks.y=element_blank(),
        axis.text.y=element_blank(),
        axis.text.x=element_text(size=7),
        axis.line = element_line(linewidth=0.3),
        legend.position = "none",
        panel.grid.major.x = element_line(colour="grey", linewidth=0.3))
```

```{r, eval=FALSE}
df_all_sub_ob_just_reordered_plot
ggsave("figures/df_all_sub_ob_just_reordered_plot.tiff", dpi=300) 
#manually remove the 3.a, 3.b. and 4. from the figure.
```

@@

```{r}
standards_conc_percentages_plot <- ggplot(totals_standards_conc, 
       aes(x=std_mcat, y=conc, fill= total_perc_in_all ))+ 
  geom_tile() + 
  scale_fill_gradient2(low = "#AFAFED", mid= "#3F3FD4", high = "#191970", midpoint = 0.1, limits = c(0,.21), na.value = "grey40") +
  scale_x_discrete(limits=(totals_standards_conc$std_mcat)[order(totals_standards_conc$group)][c(4,3,2,1)]) + 
  scale_y_discrete(limits=(totals_standards_conc$conc)[order(totals_standards_conc$group)] ) +
  geom_text(aes(label = total_perc_in_all |> percent(accuracy = 1)), color = "white", size = 2.5) +
  theme_classic() +
  coord_flip()+
  theme(legend.position = "none",
        axis.ticks = element_blank(),
        axis.text.x=element_text(angle=20,hjust=1, colour = "black"),
        axis.text = element_text(size=7, colour = "black"),
        axis.title = element_blank()) 
```

@@

```{r, warning=FALSE}
#the heatmap on top and table of stand/concept definitions on the bottom.
standards_conc_percentages_plot_and_table <- cowplot::plot_grid(
  gen_grob(totals_standards_table_flex, scaling="fixed", fit="auto"),
  standards_conc_percentages_plot,
  NULL,
  gen_grob(totals_conc_table_flex, scaling = "fixed", fit="auto"),
  nrow = 2, ncol=2, rel_heights = c(0.8,0.7), rel_widths = c(0.8,4) )
```

```{r, eval=FALSE}
standards_conc_percentages_plot_and_table
ggsave("figures/standards_conc_percentages_plot_and_table.tiff", dpi=300)
```

@@

Transportation systems, as a class of essential technologies that facilitate or impede social inclusion and activity participation<!---[@churchTransport2000; @lucasTransport2001; @socialExclusion2003; @cassSocial2005; @casasSocial2007; @prestonAccessibility2007; @paezMobility2009]-->, have increasingly come into focus from the perspective of equity; a lively and rapidly growing literature has emerged on the topic [@martensTransport2016; @karnerAdvancesPitfallsMeasuring2024; @vecchioTransportEquityLatin2020]. <!--@diciommoTransport2017; @guoSystematicOverviewTransportation2020; @karnerFrom2020; @pereiraTransportationEquity2021; @vanweeEvaluating2021; @zhangLiterature2021; @desjardinsExaminingEquityAccessibility2022; --> Aside from an ethical motivation, keeping track of objective and perceived inequalities could be of interest for governing bodies to respond to popular demands for fairness. However, transportation systems are notoriously complex: they simultaneously move people, goods, and information. Emerging technologies and service models can swiftly change the balance of benefits and burdens among a population [@guoSystematicOverviewTransportation2020], turning both users and service providers into digital rentiers [@birchBig2022]. The benefits and burdens of transportation systems are diffuse over space and time posing a challenge in evaluating and tracking inequities. Transportation systems that are engineered to offer higher mobility for people _somewhere_ can simultaneously cut others off from essential opportunities _elsewhere_, as @rajeEngineering2004 poignantly illustrated with examples of infrastructure in the UK. Over time, the shades of policies past can continue to haunt a region and the planet as shown by the legacy of displacement and decay caused by urban highways all across the US [@archerWhite2020] and the time horizon for the impacts of climate change [@markolfTransportation2019].

<!--But complexity is no excuse to shirk away from the task.--> The objective of this work is to scan the state of knowledge in terms of defining and operationalizing "fairness" in the transportation domain. Much research has been devoted to the issues of _measuring_ equity in transportation, including (among many others) [@ramjerdiEquity2006; @delboscUsing2011; @welchMeasure2013; @martensMeasuring2019; @pritchardNew2022]. Further, there are multiple works that discuss the conceptual and philosophical foundations of equity and fairness in transportation [@martensTransport2016; @pereiraDistributiveJusticeEquity2017a; @vanoutriveHow2019]. Previous reviews of planning documents have investigated equity from narrowly scoped perspectives, such as accessibility [@boisjolyHowGetThere2017] or a particular mode of transportation e.g., cycling [@doranPursuitCyclingEquity2021]. These efforts are valuable to scholars, the public, and decision-makers alike. In our estimation however, there remains a gap in the literature in terms of understanding how standards for equity are developed and implemented within the transportation domain.

To illustrate this gap, we note how @beilerExploring2016 in their exploration of transport equity, cite the following strategies identified by the U.S. Department of Transportation to address matters of justice [p. 287]:

@@

While commendable, the strategies are too vague, which means it is possible to implement them in a myriad of ways, either genuinely to comply with the spirit of justice, or else performatively to deceive it [@mcculloughPerformative2023]. Some relevant questions include: how much should the adverse effects be reduced? To zero? Or to some tolerable level of adversity greater than zero? What should that level be and by what criteria? What benefits should be distributed? How much of the benefits should minority and low-income residents receive for policymakers to declare that a project is sufficiently ‘equitable'? These questions touch on the development and use of _standards_ for transportation justice. In this work, we choose to focus exclusively on the academic domain as we know that adoption of equity concepts in planning practice has lagged developments in academic work [@pereiraTransportationEquity2021; @boisjolyHowGetThere2017; @linovskiEquityTransportationPlanning2020; @doranPursuitCyclingEquity2021; @litmanEvaluatingTransportationEquity2022]

Supporting the creation of (more) just transportation systems involves understanding the production and management of transportation benefits and costs; how they are distributed and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021]. In this spirit, our work seeks to make two significant contributions. First, it outlines a conceptual and flexible framework with supporting definitions for structuring transportation fairness and equity analysis. Second, it applies the framework to collate the existing academic knowledge that undertakes transportation equity analysis supported by equity standards and conceptualisations. We hope the findings and the presentation of academic knowledge will be useful in supporting the development and implementation of equity standards for transportation planners, policy, and decision makers. 
In contrast to previous reviews on how to measure inequality in transportation systems, this work is concerned with the implied or explicit standards that are used to judge whether the measured inequalities are fundamentally “fair” or unacceptable. The present study achieves this objective by scanning the state of academic knowledge in defining “fairness” in the transport domain. Borrowing from philosophers of justice [@jaggarPhilosophicalChallengesGlobal2009], we ask the following questions to the literature and use them as a framework for presentation:

-	**What** are those benefits and burdens of transportation systems? Are they qualified in sidewalk widths, bus frequency, travel times, accessibility, or some other consideration?  
-	**Who** benefits and who is burdened by transportation systems? Along which lines of identity are transportation inequities examined?
-	**Where and when** are equity standards applied in the literature? What are the differences between the research depending on location, context-specific planning processes and time of publication? 
-	**How** are equity standards and inequities measured? Are they infrastructure provision thresholds, acceptable pollution levels, or some other kind of cut off? How are these studies born out of broader notions of justice, i.e. disability rights, environmental justice, egalitarianism, or something else?

@@

In the introductory section the terms "justice", "equity", and "fairness" are used relatively loosely. This was done purposefully. As noted, people often have strong intuitions of what is "just", "equitable", and "fair" which may or may not match those of the authorities who set the "standards" of fairness. But in democratic societies, the authority of all those charged with the business of governing should derive from the will of the people. It is therefore important to explicitly state what we mean by these terms. In the following, we describe justice as embracing moral motivation, equity as measurement tools, fairness as a yardstick of justice, and standards as statements of fairness.

@@

- **Retributive justice** is concerned with the proportional retribution of wrongdoers with relation to legitimate punishers and the innocent [@sep-justice-retributive].
- **Reparative (or restorative) justice** focuses on the reparation of caused harm; it centers the needs and voices of victims to restore wrongdoers and the community according to moral values [@tylerRestorativeJusticeProcedural2006; @braithwaite2003principles]. In planning and policy contexts, reparative justice is often conceived of as accountability mechanisms in which institutions that caused harm financially compensate victims [@safransky2022grammars]. 
- **Procedural justice** strives to ensure that the views and preferences of all stakeholders are fairly accounted for in the decision-making and inter-personal procedures affecting their lives and communities [@tylerRestorativeJusticeProcedural2006].
- **Distributive justice** is perhaps the most studied form of justice [see @jaggarPhilosophicalChallengesGlobal2009, p. 2] as well in the transportation domain [@pereiraDistributiveJusticeEquity2017a]. Its main concern is the way the benefits and burdens of the tangible and intangible products of society are collected by different segments of a population.

@@

**Equity**, as conceptualized alongside distributive justice, tends to encompass tools to understand the distribution of benefits and burdens of things among a population, often emphasizing those with the least advantage and their outcomes. In the transportation domain, the term is somewhat loaded because it is perceived as stemming from the authority of the state and is meant to assist with decisions about regulating and financing transportation spending [@karnerTransportationEquityTransportation2020]. Here, we are in agreement with @karnerTransportationEquityTransportation2020 that equity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual and perceived inequities. In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in the "political ecology of the estimated truth" [@kingModels1993]: in principle, their assumptions and scope must be open and transparent, or else they may be more vulnerable to misuse and even abuse as tools of subjugation.

@@

**Standards** are a way of making concrete statements about fairness. Returning to the ambiguities in Title VI discussed by @martensFair2021, the attainment of justice depends on the standard used to indicate fairness. For example, explicit non-discrimination constitutes a very weak standard that takes aim at the actions of agencies instead of the recipients of the benefits. Accordingly, any distribution of benefits would be considered fair, as long as the agency does not explicitly and knowingly target or deny benefits to particular groups. The standard provides conditions to determine whether a situation is _fair_. A similarly weak standard is a Pareto improvement, whereby it is possible to concentrate the benefits as long as no group is worse off compared to the status quo e.g., [@tan2016pareto; @xuParetoimprovingPoliciesIdealized2018]. From this standard, a policy could be considered fair as long it does no harm. <!--A somewhat more strict standard, a Pareto-Plus improvement, stipulates that an intervention is fair when all groups receive at least _some_ (non-trivial) benefits; the size of the benefits for each group is irrelevant. In contrast to the notion of "do no harm", such a standard embodies the ideal that no one is denied benefits. --> A standard based on egalitarian theories such as proportional equity would weigh the benefits or burdens by population, and fairness is achieved when each group gives or receives in proportion to their size [@billsLookingMeanEquity2017;@martensJusticetheoreticApproachDistribution2012] In contrast, an affirmative action standard based in restorative justice is even stricter. It requires the benefits to be distributed in a non-egalitarian way that favors those who are still harmed by past or present discriminatory practices [@bierbaumMobilityJusticeLinking2021].

@@

In the case of justice in transportation, the question of **"Where?"** is paramount, as it might be argued that transportation generates inequalities by their very nature. By concentrating the effects of space-time convergence (for instance, by providing access to a transit system or a highway), an inequality is automatically generated. The burdens of transportation, in contrast, are often diffuse. They are incrementally paid, for example by a distributed population in the form of taxes, or by a population with a different spatial distribution in the form of poor health. As such, the answer to "Where?" is the definition of the spatial boundaries. For example, we can distinguish literature from the global South and North, allowing us to examine how they approach, deal with, and position themselves in different spatial contexts in addressing transport inequalities.

Conventionally, the question of **"When?"** refers to the temporal circumstances within which the demands of justice have application. In the cause of transportation justice, one could ask about the temporal aspects of transportation systems, as examples: _when_ did the equity analysis take place and under what historical policy context; _the right time_ to invest in transportation infrastructure and as a result when to generate a spatial inequality [@rabelloEvaluation2015]; for _how long_ the burdens and benefits can still be associated to a specific transportation intervention; or even _timelines_ of reparative justice interventions that reconcile the shadows of past transportation-related injustices. In this literature review, given the amount of papers reviewed and the inconsistencies in how temporal aspects were addressed, we exclusively focus on the paper's publication year.

When answering **"Who?"**, we inquire about which entities should be regarded as subjects/arbiters of justice, meaning those entitled to make claims of moral consideration from the perspective of justice. To make it tractable, this question is often approached through the filter of population groups, which may include several concurrent traits, such as gender identity, ableness, ethnicity, age, caste, and income. Often, it is appropriate to reflect on the intersections between traits, given differences in a person's lived experiences. A possible complication in the case of transportation is that disentangling the "Who" from their mobility tools is not always straightforward. Although a person is not their mode of transportation, there are large segments of the population who live in situations where they cannot extricate themselves from the mobility tools they can use, either because they have driven themselves out of choices [@laveryDriving2013], or have been driven out of choices by factors beyond their control such as captive users of a single mode [@jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018]. <!-- In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume.-->While it is important to avoid conflating the "Who" with the "What", for analytical purposes we need to be mindful of the connection between a person and their mobility tools. <!-- In the case of transportation, in addition to members of the public who use transportation systems, there is another category of **Who**, that stands possibly in opposition to users, namely the entities charged with providing services, maintaining infrastructure, and so on. These could be ministries or departments of transportation, transit agencies, public works departments and others having the power to act upon transportation (in)justices. Identifying these entities is relevant to elucidate who is responsible, for example for apportioning the benefits or mitigating the burdens of transportation.-->
